How to select predictive models for decision-making or causal inference?

### 2023-12-08 how_to_select_causal_models.tex
move definitions to appendix

\affil[1]{Inria, Soda, Saclay, France}
\affil[2]{Mission Data, Haute Autorité de Santé, Saint-Denis, France}
\affil[*]{Corresponding author: m.doutreligne@has-sante.fr}

@@

    \caption[The best predictor may not estimate best causal
        effects]{\textbf{Illustration}: a) a random-forest estimator
        with high performance for standard prediction (high $R^2$) but that
        yields poor causal estimates (large error between true effect $\tau$ and
        estimated $\hat{\tau}$), b) a linear estimator with smaller
        prediction performance leading to better causal estimation. \\[1ex]
        Selecting the estimator with the smallest error to the individual
        treatment effect $\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$
        --the $\tau\text{-risk}$, def.\,\ref{def:tau_risk} -- would lead to
        the best causal estimates; however computing this error is not
        feasible: it requires access to unknown quantities:
        $\tau(x)$. \\[1ex]
        While the random forest fits the data better than the linear model, it
        gives worse causal inference because its error is inhomogeneous between
        treated and untreated. The $R^2$ score does not capture this
        inhomogeneity.
        \label{fig:toy_example}
    }
\end{figure}

@@

%\section{Formal setting of causal inference and model selection}\label{sec:framework}
\section{Methods}\label{sec:framework}
\subsection{The Neyman-Rubin Potential Outcomes framework}%
\label{sec:neyman_rubin}%

@@

The Neyman-Rubin Potential Outcomes framework
\cite{naimi2023defining,imbens_causal_2015} enables statistical reasoning on
causal treatment effects: Given an outcome $Y \in \mathbb R$ (eg. mortality risk
or hospitalization length), function of a binary treatment $A \in \mathcal{A} =
    \{0, 1\}$ (eg.~a medical procedure, a drug administration), and baseline
covariates $X \in \mathcal{X} \subset \mathbb{R}^d$, we observe the factual
distribution, $O = (Y(A), X, A) \sim \mathcal D = \mathbb P(y, x, a)$. However,
we want to model the existence of potential observations (unobserved ie.
counterfactual) that correspond to a different treatment. Thus we want
quantities on the counterfactual distribution $O^{*} = (Y(1), Y(0), X, A) \sim
    \mathcal D^{*} = \mathbb P(y(1), y(0), x, a)$.

@@

Should we know the two expected outcomes for a given $X$,
we could compute the
difference between them, which gives the causal effect of the treatment.
%
These two expected outcomes can be computed from the observed data:
the consistency \ref{assumption:consistency} and unconfoundedness
\ref{assumption:ignorability} assumptions imply the equality of two different
expectations:
\begin{equation}\label{eq:mu_identification}
    \mathbb E_{Y(a) \sim \mathcal{D^{\star}}} [Y(a)|X=x] = \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
\end{equation}
On the left, the expectation is taken on the counterfactual unobserved
distribution. On the right, the expectation is taken on the factual observed
distribution conditionally on the treatment. For the rest of the
paper, the expectations will always be taken on the factual observed
distribution $\mathcal{D}$. This identification leads to outcome based estimators (ie.
g-computation estimators \cite{snowden_implementation_2011}):
\begin{eqnarray}
    \tau =& \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)|X=x]
    \notag
    \\
    = &\mathbb E_{Y \sim \mathcal{D}}[Y|A=1] - \mathbb E_{Y \sim \mathcal{D}}[Y| A=0]
    \label{eq:tau_population}
\end{eqnarray}
This equation builds on two quantities: the conditional expectancy
of the outcome given the covariates and either
treatment or no no treatment, called \emph{response function}:
\begin{flalign*}
    \text{Response function}
     &  &
    \mu_{a}(x) \myeq \; \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
     &  &
\end{flalign*}

@@

\paragraph{The $\tau\text{-risk}$: an oracle error risk}\label{paragraph:oracle_metrics}
%
As we would like to target the CATE, the following
evaluation risk is natural:
\begin{definition}[$\tau\text{-risk}(f)$]\label{def:tau_risk} also called PEHE
    \cite{schulam_reliable_2017, hill_bayesian_2011}:
    \begin{equation*}\label{eq:tau_risk}
        \tau\text{-risk}(f) = \mathbb E_{X\sim p(X)}[(\tau(X) - \hat \tau_f(X))^2]
    \end{equation*}
\end{definition}

@@

\paragraph{Feasible error risks}\label{paragraph:feasible_metrics}
\emph{Feasible} risks are based on the prediction error of the outcome model
and \emph{observable} quantities.

All expectations below are on observed distribution:
$(Y, X, A) \sim \mathcal D$.

We now detail risks that use the nuisances $e$
--propensity score, def \ref{def:propensity_score}-- and $m$ --conditional mean
outcome, def \ref{def:m}. We give the definitions as \textit{semi-oracles},
function of the true unknown nuisances, but later instantiate them with estimated
nuisances, noted $\big(\check e, \check m \big)$. Semi-oracles risks are
superscripted with the $^{\star}$ symbol.

These risks are summarized in Table \ref{tab:evaluation_metrics}.

@@

Causal model selection (as in
\autoref{eq:causal_model_selection}) may involve estimating various quantities
from the observed data: the outcome model $f$, its induced risk as
introduce in the previous section, and possibly nuisances required by the
risk.
Given a dataset with $N$ samples, we split out a train and a test sets
$(\mathcal{T}, \mathcal{S})$. We
fit each candidate estimator $f \in \mathcal{F}$ on $\mathcal{T}$. We also fit
the nuisance models $(\check e, \check m)$ on the train set
$\mathcal{T}$, setting hyperparameters by a nested
cross-validation before fitting the nuisance estimators with these parameters
on the full train set. Causal quantities are then computed by applying the fitted  candidates
estimators $f \in \mathcal{F}$ on the test set $\mathcal{S}$. Finally, we
compute the model-selection metrics for
each candidate model on the test set. This procedure is described in Algorithm
\ref{problem:estimation_procedure:algo} and Figure
\ref{problem:estimation_procedure:figure}.

@@

\section{A theoretical argument favoring the R-risk}\label{theory:r_risk_rewrite}

We now derive a novel rewriting of the $R\text{-risk}$, making explicit its link
with the oracle $\tau \text{-risk}$.

@@

%\idea{Overlap is important but not easily measurable}
Good overlap between treated and control population is
crucial for causal inference as it is required by the positivity assumption
\ref{assumption:overlap}. It is often assessed by
comparing visually population distributions (as in Figure
\ref{fig:toy_example}) or computing standardized difference on each
feature
\cite{austin_introduction_2011,austin_moving_2015}. While these methods are
useful to decide if positivity holds, they do not yield a
single measure. Rather, we compute the divergence
between the population covariate distributions $\mathbb P(X|A=0)$ and $\mathbb
    P(X|A=1)$
\cite{damour_overlap_2020,johansson2022generalization}. We introduce the
Normalized Total Variation (NTV), a divergence based on the sole propensity
score (see \ref{apd:motivation_ntv}).

@@

Figure \ref{fig:relative_kendalls_all_datasets} shows the agreement between the
ideal ranking of outcome models given the oracle $\tau\text{-risk}$ and the different
feasible causal metrics. We measure this agreement with a relative\footnote{To
    remove the variance across datasets (some datasets lead to easier model
    selection than others), we report values for one metric relative to the mean of
    all metrics for a given dataset instance: $\text{Relative} \,
        \kappa(\ell,\tau\mathrm{{-risk}})= \kappa(\ell,\tau\mathrm{{-risk}}) -
        mean_{\ell}\big(\kappa(\ell,\tau\mathrm{{-risk}})\big)$} Kendall tau $\kappa$
(eq. \ref{eq:kendall_tau}) \cite{kendall_new_1938}. Given the importance of
overlap in how well metrics approximate the oracle $\tau\text{-risk}$, we
separate strong and weak overlap.

@@

Note that propensity score models must be selected to estimate
the individual posterior probability. For this, we used the Brier score,
is minimized by the true individual probability. An easy mistake is to
use calibration errors popular in machine learning
\cite{platt_probabilistic_1999,zadrozny_obtaining_2001,niculescu-mizil_predicting_2005,minderer_revisiting_2021}
as these select not for the individual posterior probability but for an
aggregate error rate \cite{perez2022beyond}.

\paragraph{Extension to binary outcomes}
While we focused on continuous outcomes, in medicine, the target outcome
is often a categorical variable such as mortality status or diagnosis. In
this case, it may be interesting to focus on other estimands than the
Average Treatment Effect $\mathbb{E}[Y(1)] -\mathbb{E}[Y(0)] $, for instance the relative risk
$\frac{\mathbb P(Y(1) = 1)}{\mathbb P(Y(0) = 1)}$
or the odd ratio, $\frac{\mathbb P(Y(1) = 1) / [1 - \mathbb P(Y(1)
    =1)]}{\mathbb P(Y(0) = 1) / [1 - \mathbb P(Y(0) = 1]}$ are often used
\cite{austin2017estimating}. While the odds ratio is natural for
case-control studies \cite{rothman2008case}, other measures
can reduce heterogeneity \cite{colnet2023risk}.
In the log domain, the ratios are written as a difference, the
framework studied here (\autoref{sec:neyman_rubin}) can directly apply.
In particular, the log odds ratio is estimated by the common
cross-entropy loss (or log loss) as in logistic regression.

@@

>>>>

\affil[1]{Inria, Soda, Saclay, France}
\affil[2]{Mission Data, Haute Autorité de Santé, Saint-Denis, France}
\affil[*]{Corresponding author: matt.dout@gmail.com}

@@

    \caption[The best predictor may not estimate best causal
        effects]{\textbf{Illustration}: a) a random-forest estimator
        with high performance for standard prediction (high $R^2$) but that
        yields poor causal estimates (large error between true effect $\tau$ and
        estimated $\hat{\tau}$), b) a linear estimator with smaller
        prediction performance leading to better causal estimation. \\[1ex]
        Selecting the estimator with the smallest error to the individual
        treatment effect $\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$
        --the $\tau\text{-risk}$, eq.\,\ref{eq:tau_risk} -- would lead to
        the best causal estimates; however computing this error is not
        feasible: it requires access to unknown quantities:
        $\tau(x)$. \\[1ex]
        While the random forest fits the data better than the linear model, it
        gives worse causal inference because its error is inhomogeneous between
        treated and untreated. The $R^2$ score does not capture this
        inhomogeneity.
        \label{fig:toy_example}
    }
\end{figure}

@@

%\section{Formal setting of causal inference and model selection}\label{sec:framework}
\section{Methods}\label{sec:framework}
\subsection{Neyman-Rubin Potential Outcomes framework}%
\label{sec:neyman_rubin}%

@@

The Neyman-Rubin Potential Outcomes framework
\cite{naimi2023defining,imbens_causal_2015} enables statistical reasoning on
causal treatment effects: Given an outcome $Y \in \mathbb R$ (eg. mortality risk
or hospitalization length), function of a binary treatment $A \in \mathcal{A} =
    \{0, 1\}$ (eg.~a medical procedure), and baseline
covariates $X \in \mathcal{X} \subset \mathbb{R}^d$, we observe the factual
distribution, $O = (Y(A), X, A) \sim \mathcal D = \mathbb P(y, x, a)$. However,
we want to model the existence of potential observations (unobserved ie.
counterfactual) that correspond to a different treatment. Thus we want
quantities on the counterfactual distribution $O^{*} = (Y(1), Y(0), X, A) \sim
    \mathcal D^{*} = \mathbb P(y(1), y(0), x, a)$.

@@

Should we know the two expected outcomes for a given $X$,
we could compute the
difference between them, which gives the causal effect of the treatment.
%
These two expected outcomes can be estimated from observed data:
the consistency \ref{assumption:consistency} and unconfoundedness
\ref{assumption:ignorability} assumptions imply the following equality:
\begin{equation}\label{eq:mu_identification}
    \mathbb E_{Y(a) \sim \mathcal{D^{\star}}} [Y(a)|X=x] = \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
\end{equation}
On the left, the expectation is taken on the counterfactual unobserved
distribution. On the right, the expectation is taken on the factual observed
distribution conditionally on the treatment. For the rest of the
paper, the expectations will always be taken on the factual observed
distribution $\mathcal{D}$. This identification leads to outcome based estimators (ie.
g-computation estimators \cite{snowden_implementation_2011}):
\begin{eqnarray}
    \tau =& \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)|X=x]
    \notag
    \\
    = &\mathbb E_{Y \sim \mathcal{D}}[Y|A=1] - \mathbb E_{Y \sim \mathcal{D}}[Y| A=0]
    \label{eq:tau_population}
\end{eqnarray}
This equation builds on two quantities: the conditional expectancy
of the outcome given the covariates and either
treatment or no no treatment, called \emph{response function}:
\begin{flalign*}
    \text{Response function}
     &  &
    \mu_{a}(x) \myeq \; \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
     &  &
\end{flalign*}

@@

\paragraph{The $\tau\text{-risk}$: an oracle error risk}\label{paragraph:oracle_metrics}
%
As we would like to target the CATE, the following
evaluation risk is natural (also called PEHE \cite{schulam_reliable_2017, hill_bayesian_2011}):
\begin{flalign}\label{eq:tau_risk}
    \tau\text{-risk}(f) \myeq \mathbb E_{X\sim p(X)}[(\tau(X) - \hat \tau_f(X))^2]
\end{flalign}

@@

\paragraph{Feasible error risks}\label{paragraph:feasible_metrics} Table
\ref{tab:evaluation_metrics} lists \emph{feasible} risks (Detailed in Appendix
\ref{def:feasible_risks}), based on the prediction error of the outcome model
and \emph{observable} quantities. These observable, called nuisances are $e$
--propensity score, eq \ref{def:propensity_score}-- and $m$ --conditional mean
outcome, eq \ref{def:m}. We give the definitions as \textit{semi-oracles},
function of the true unknown nuisances, but later instantiate them with
estimated nuisances, noted $\big(\check e, \check m \big)$. Semi-oracles risks
are superscripted with the $^{\star}$ symbol.

@@

Causal model selection (eq
\ref{eq:causal_model_selection}) may involve estimating various quantities
from the observed data: the outcome model $f$, its induced risk as
introduce in the previous section, and possibly nuisances required by the
risk.
Given a dataset with $N$ samples, we split out a train and a test sets
$(\mathcal{T}, \mathcal{S})$. We
fit each candidate estimator $f \in \mathcal{F}$ on $\mathcal{T}$. We also fit
the nuisance models $(\check e, \check m)$ on the train set
$\mathcal{T}$, setting hyperparameters by a nested
cross-validation before fitting the nuisance estimators with these parameters
on the full train set. Causal quantities are then computed by applying the fitted  candidates
estimators $f \in \mathcal{F}$ on the test set $\mathcal{S}$. Finally, we
compute the model-selection metrics for
each candidate model on the test set. This procedure is described in Algorithm
\ref{problem:estimation_procedure:algo} and Figure
\ref{problem:estimation_procedure:figure}.

@@

\section{The R-risk, a reweighted oracle metric}\label{theory:r_risk_rewrite}

The $R\text{-risk}$ can be rewritten as a rebalanced $\tau \text{-risk}$.

@@

%\idea{Overlap is important but not easily measurable}
Good overlap between treated and control population is
crucial for causal inference (positivity assumption
\ref{assumption:overlap}). It is often assessed by
comparing visually population distributions as in Figure
\ref{fig:toy_example} or computing standardized difference on each
feature
\cite{austin_introduction_2011,austin_moving_2015}. While these methods are
useful to decide if positivity holds, they do not yield a
single measure. Rather, we compute the divergence
between the population covariate distributions $\mathbb P(X|A=0)$ and $\mathbb
    P(X|A=1)$
\cite{damour_overlap_2020,johansson2022generalization}. We introduce the
Normalized Total Variation (NTV), a divergence based on the sole propensity
score (Appendix \ref{apd:motivation_ntv}).

@@

Figure \ref{fig:relative_kendalls_all_datasets} shows the agreement between the
ideal ranking of outcome models given the oracle $\tau\text{-risk}$ and the different
feasible causal metrics. We measure this agreement with relative\footnote{To
    remove the variance across datasets (some datasets lead to easier model
    selection than others), we report values for one metric relative to the mean of
    all metrics for a given dataset instance: $\text{Relative} \,
        \kappa(\ell,\tau\mathrm{{-risk}})= \kappa(\ell,\tau\mathrm{{-risk}}) -
        mean_{\ell}\big(\kappa(\ell,\tau\mathrm{{-risk}})\big)$} Kendall tau $\kappa$
(eq. \ref{eq:kendall_tau}) \cite{kendall_new_1938}. Given the importance of
overlap in how well metrics approximate the oracle $\tau\text{-risk}$, we
separate strong and weak overlap.

@@

Note that propensity score models must be selected to estimate
the individual posterior probability. For this, we used the Brier score,
which is minimized by the true individual probability. An easy mistake is to
use calibration errors popular in machine learning
\cite{platt_probabilistic_1999,zadrozny_obtaining_2001,niculescu-mizil_predicting_2005,minderer_revisiting_2021}
as these select not for the individual posterior probability but for an
aggregate error rate \cite{perez2022beyond}.

\paragraph{Extension to binary outcomes}
While we focused on continuous outcomes, in medicine, the target outcome
is often a categorical variable such as mortality status or diagnosis. In
this case, it may be interesting to focus on other estimands than the
Average Treatment Effect $\mathbb{E}[Y(1)] -\mathbb{E}[Y(0)] $, for instance the relative risk
$\frac{\mathbb P(Y(1) = 1)}{\mathbb P(Y(0) = 1)}$
or the odd ratio, $\frac{\mathbb P(Y(1) = 1) / [1 - \mathbb P(Y(1)
    =1)]}{\mathbb P(Y(0) = 1) / [1 - \mathbb P(Y(0) = 1]}$ are often used
\cite{austin2017estimating}. While the odds ratio is natural for
case-control studies \cite{rothman2008case}, other measures
can reduce heterogeneity \cite{colnet2023risk}.
In the log domain, the ratios are written as a difference, the
framework studied here can directly apply.
In particular, the log odds ratio is estimated by the common
cross-entropy loss (or log loss) as in logistic regression.

@@

\section{Definitions of feasible risks}\label{def:feasible_risks}
